---
title: Tidy Tuesday pt 1
author: Alexander C. Hungerford
date: '2020-12-06'
slug: []
categories:
  - Tidy Tuesday
tags:
  - dataViz
  - ggplot2
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---


### Setup

```{r}
library(tidyverse)
```

### Load data

```{r}
mtcars_tbl <- tibble(mtcars)
mtcars_tbl
```

```{r}
# basic scatterplot
ggplot(mtcars_tbl, aes(x = cyl, y = mpg)) + geom_point()
```

The x-axis is a discrete variable. The number of cylinders in a vehicle engine is always an integer (1, 2, 3...n). You will never have a car with 4.25 cylinders.

As we can see, generally, vehicles with more cylinders get worse gas mileage.

### Code Explained

```{r, include=FALSE}
# basic scatterplot
ggplot(mtcars_tbl, aes(x = cyl, y = mpg)) + geom_point()
```

`ggplot()` calls the **ggplot** package and initializes a ggplot object. In the first argument you need to put the data you want to work from, `mtcars_tbl`. The second argument is the aesthetics layer. `aes(x = cyl, y = mpg)` can be written `aes(cyl, mpg)`. The x and y are understood implicitly, but for others reading your code it helps to be as explicit as possible.

The final layer is the **geom** layer. This tells ggplot what type of plot to make. If you had `geom_bar` would make a bar plot. `geom_histogram` makes a histogram and so on.

```{r, include=FALSE}
# this would produce a plot identical to the scatterplot above
# but it is less explicit and more difficult for others to understand your code
ggplot(mtcars_tbl, aes(cyl, mpg)) + geom_point()
```

### Aesthetic Layer

Aesthetics describe how variables are mapped to visual properties. You can do very simple manipulations in the aesthetic layer, such as `log(mpg)`, but if you need to do anything complicated you should do another layer and `mutate()` the data.

```{r}
# mpg by weight, data color by cylinder
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) + geom_point()
```

```{r}
# a simple manipulation of the data in the aes() layer with log(wt)
ggplot(mtcars, aes(x = log(wt), y = mpg, color = cyl)) + geom_point()
```

You can also map aesthetics to functions of variables.

```{r}
# x-axis is horsepower per cylinder
ggplot(mtcars, aes(x = hp/cyl, y = mpg, color = cyl)) + geom_point()
```

For most 